---
title: "chronoepigenetic oscillations"
output: html_document
---

## Setup
```{r}
require(ggplot2)
require(dplyr)
require(data.table)
require(reshape2)
require(png)
coef2wave <- function(amp,acro,int=0,zts=0:24,per=24){return(amp/2*cos(zts/24*2*pi-acro/24*2*pi)+int)}


```

## Idiosyncratic desynchronosis

### Figure 5a
```{r fig.height= 4, fig.width=3}

res_list <- vector('list', 50)
phase_shift = 2

for(i in 1:length(res_list)){
  hypothetical_amplitude <- 5
  hypo_amp <- sample(seq(hypothetical_amplitude/2, hypothetical_amplitude, hypothetical_amplitude/10),1)
  # hypo_amp <- 5
  tmp2 <- data.frame(rep=i, zt = 0:24, 
                     healthy = coef2wave(amp=hypo_amp, int= 2.5,
                                                 acro=sample(c(seq(21, 24, 0.5), seq(0, 3, 0.5)), 1)), 
                     affected = coef2wave(amp=hypo_amp, int= 2.5,
                                                  acro=sample(c(seq(21-phase_shift[1], 24, 0.5), seq(0, 3+phase_shift[1], 0.5)), 1))) 
  res_list[[i]] <- tmp2
  }

res <- do.call(rbind, res_list)
res <- res %>%
  melt(., id.var = c("rep","zt")) 

graph1 <- res %>%
  ggplot(aes(x=zt, y=value, col=as.factor(rep))) +
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "none") +
  ylab("normalized\ncytosine modification\ndensity (%)") + xlab("time of day") +
  scale_x_continuous(breaks=seq(0, 24, 4), labels = c("12AM","4AM","8AM","12PM","4PM","8PM","12AM")) +
  geom_rect(aes(xmin=9,xmax=17,ymin=-Inf,ymax=Inf), alpha=0.1, fill="grey90", col="white") +
  geom_smooth(method="lm", size=0.25, formula= y ~ sin(x/24*2*pi) + cos(x/24*2*pi), se=FALSE) +
  facet_wrap(~variable, nrow=2) + scale_color_manual(values=c(rep("black", 50))) 
  # facet_wrap(~variable) + scale_color_manual(values=c(viridis::cividis(20))) 

graph1

### Save output
# saveRDS(res, "output/chrono_simulation.RDS")

### Load published data
 res <- readRDS("input/chrono_simulation.RDS")

```

### Figure 5b: sampling simulations
```{r fig.height= 4, fig.width=3}

sample_points <- res %>%
  filter( zt > 9, zt < 17 ) %>%
  dcast(rep + zt ~ variable, value.var = "value", sum) %>%
  group_by(rep) %>%
  sample_n(., 1) %>%
  melt(., id.var = c("rep","zt")) 

graph1 +
  geom_point(data= sample_points, aes(x=zt, y=value), col="darkred", alpha=0.75)

### Save output
# saveRDS(sample_points, "output/chrono_samplepoints.RDS")

### Load published data
sample_points <- readRDS("input/chrono_samplepoints.RDS")

```

### Figure 5c: sampled boxplot
```{r fig.height= 4, fig.width=3}

sample_points %>%
  ggplot(aes(x=variable, y= value)) +
  geom_boxplot() +
  geom_dotplot(binaxis='y', stackdir='center', col = "white", fill="darkred", stackratio=1, dotsize=1) + 
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "none") +
  ylim(c(0,5)) +
  ylab("cytosine modification density \n at work-hour sampling") + xlab("co-twin cohort")
  # stat_summary(fun=median, geom="point", shape=18, size=3, color="black") + 
  # stat_summary(fun.data="mean_sdl", fun.args = list(mult=1),geom="crossbar", shape=18, col="grey25", width=0.5)

```

### Statistical tests
```{r}
df1 <- sample_points %>%
  dcast(rep ~ variable, value.var = "value", sum)

tres <- t.test(df1$healthy, df1$affected, alternative="two.sided", var.equal=FALSE)
bres <- bartlett.test(data = sample_points, value ~ variable)
ares <- aov(data = sample_points, value ~ variable)
summary(ares)[[1]][["Pr(>F)"]][1]
tres$p.value
bres$p.value

```

<!-- ### Sampling test with label -->
<!-- ```{r fig.height= 4, fig.width=3} -->
<!-- text_values <- paste("t-test p=", round(tpval, 2), -->
<!--       "\nanova p=", round(apval, 2), -->
<!--       "\nbartlett's p=", round(bpval, 2), sep="") -->

<!-- sample_points %>% -->
<!--   ggplot(aes(x=variable, y= value)) + -->
<!--   geom_boxplot() + -->
<!--   geom_dotplot(binaxis='y', stackdir='center', col = "white", fill="darkred", stackratio=1, dotsize=1) +  -->
<!--   theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "none") + -->
<!--   ylab("cytosine modification density \n at work-hour sampling") + xlab("co-twin cohort") + -->
<!--   ylim(c(0,5)) + -->
<!--   annotate("text", x=Inf, y=Inf, label = text_values, vjust=1, hjust=1) -->

<!-- ``` -->

### Figure 5d: iterations of desynchronosis

```{r}

simulator <- function(iter){
phase_shift = seq(0, 6, 0.5)
sample_size = 50
res_list2 <- vector('list', length(sample_size))

for(j in 1:length(phase_shift)){

res_list <- vector('list', sample_size)

for(i in 1:length(res_list)){
  hypothetical_amplitude <- 5
  hypo_amp <- sample(seq(hypothetical_amplitude/2, hypothetical_amplitude, hypothetical_amplitude/10),1)
  # hypo_amp <- tmp$hyp_amp 

  tmp2 <- data.frame(rep=i, zt = 0:24, 
                     healthy_cotwins = coef2wave(amp=hypo_amp, 
                                                 acro=sample(c(seq(21, 24, 0.5), seq(0, 3, 0.5)), 1)), 
                     affected_cotwins = coef2wave(amp=hypo_amp, 
                                                  acro=sample(c(seq(21-phase_shift[j], 24, 0.5), seq(0, 3+phase_shift[j], 0.5)), 1))) 
  res_list[[i]] <- tmp2
  }

res <- do.call(rbind, res_list)
res <- res %>%
  melt(., id.var = c("rep","zt")) %>%
  mutate(value = value + abs(min(value)))

sample_points <- res %>%
  filter( zt > 9, zt < 17 ) %>%
  dcast(rep + zt ~ variable, value.var = "value", sum) %>%
  group_by(rep) %>%
  sample_n(., 1) %>%
  melt(., id.var = c("rep","zt")) 

df1 <- sample_points %>%
  dcast(rep ~ variable, value.var = "value", sum)

tres <- t.test(df1$healthy_cotwins, df1$affected_cotwins, alternative="two.sided")
bres <- bartlett.test(data = sample_points, value ~ variable)
ares <- aov(data = sample_points, value ~ variable)
res_list2[[j]] <- data.frame(phase_shift = phase_shift[j], 
           healthy_mean = mean(df1$healthy_cotwins),
           affected_mean = mean(df1$affected_cotwins),
           healthy_sd = sd(df1$healthy_cotwins),
           affected_sd = sd(df1$affected_cotwins),
           pval_ttest = tres$p.value,
           pval_anova = summary(ares)[[1]][["Pr(>F)"]][1],
           pval_bartlett = bres$p.value)
}

return(data.frame(iter=iter, do.call(rbind, res_list2)))
}

simulation <- do.call(rbind, lapply(1:1000, function(x) simulator(x)))

### Save output
# saveRDS(simulation, "output/chrono_iterations.RDS")

### Load published data
simulation <- readRDS("input/chrono_iterations.RDS")

```

#### Figure 5d cont.. pie chart
```{r fig.height= 4, fig.width=3}
colorpalette = viridis::viridis(20)

simulation %>%
  select_if(., grepl("iter|phase_shift|pval", colnames(.))) %>%
  mutate(legend = case_when(pval_bartlett < 0.05 & pval_ttest < 0.05 ~ "DMP & DVP",
                         pval_bartlett < 0.05 ~ "DVP",
                         pval_ttest < 0.05 ~ "DMP",
                         pval_bartlett > 0.05 & pval_ttest > 0.05 ~ "no significance")) %>%
  select(iter, phase_shift, legend) %>%
  group_by(phase_shift) %>%
  count(legend) %>%
  ggplot(aes(x="", y=n, fill=legend)) +
    geom_bar(stat = "identity", col="white", position = position_fill()) +
  scale_fill_manual(values=c(colorpalette[1],colorpalette[18],colorpalette[10],"grey80")) +
    coord_polar(theta = "y") +
    facet_wrap(~ phase_shift, nrow=3) +
  # xlim(c(2, 4)) +
  theme_bw() + theme(panel.grid.major = element_blank(), axis.ticks = element_blank(), axis.text.x=element_blank(), panel.grid.minor = element_blank()) +
   theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) + 
    theme(legend.position='bottom') +guides(fill=guide_legend(nrow=2,byrow=TRUE))

```


### Figure 5e: sampling time heterogeneity

```{r fig.height= 3, fig.width=3}
sample_points1 <- res %>%
    filter(variable !="healthy") %>%
  filter( zt >= 6, zt < 9 ) %>%
  dcast(rep + zt ~ variable, value.var = "value", sum) %>%
  group_by(rep) %>%
  sample_n(., 1) %>%
  melt(., id.var = c("rep","zt")) %>%
  mutate(tmp ="early\nmorning")

sample_points2 <- res %>%
    filter(variable !="healthy") %>%
  filter( zt >= 9, zt < 12 ) %>%
  dcast(rep + zt ~ variable, value.var = "value", sum) %>%
  group_by(rep) %>%
  sample_n(., 1) %>%
  melt(., id.var = c("rep","zt")) %>%
  mutate(tmp ="morning")

sample_points3 <- res %>%
    filter(variable !="healthy") %>%
  filter( zt >= 12, zt < 15 ) %>%
  dcast(rep + zt ~ variable, value.var = "value", sum) %>%
  group_by(rep) %>%
  sample_n(., 1) %>%
  melt(., id.var = c("rep","zt")) %>%
  mutate(tmp ="early\nafternoon")

sample_points4 <- res %>%
    filter(variable !="healthy") %>%
  filter( zt >= 15, zt < 18 ) %>%
  dcast(rep + zt ~ variable, value.var = "value", sum) %>%
  group_by(rep) %>%
  sample_n(., 1) %>%
  melt(., id.var = c("rep","zt")) %>%
  mutate(tmp ="late\nafternoon")

sample_points5 <- res %>%
    filter(variable !="healthy") %>%
  filter( zt >= 9, zt <= 17 ) %>%
  dcast(rep + zt ~ variable, value.var = "value", sum) %>%
  group_by(rep) %>%
  sample_n(., 1) %>%
  melt(., id.var = c("rep","zt")) %>%
  mutate(tmp ="random\nduring work-hours")


```

#### Figure 5e cont.. waveforms
```{r fig.height= 3, fig.width=5}

sam_total <- rbind(sample_points1,sample_points2,sample_points3,sample_points4, sample_points5) 

### Output data
# saveRDS(sam_total, "output/sam_total.RDS")
### Import published data
sample_points <- readRDS("input/sam_total.RDS")


colorpalette = viridis::viridis(20)

res %>%
  filter(variable !="healthy") %>%
  ggplot(aes(x=zt, y=value, col=as.factor(rep))) +
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "none") +
  ylab("normalized\ncytosine modification\ndensity (%)") + xlab("time of day") +
  scale_x_continuous(breaks=seq(0,24, 4), labels = c("12AM","4AM","8AM","12PM","4PM","8PM","12AM")) +
  # geom_rect(aes(xmin=9,xmax=17,ymin=-Inf,ymax=Inf), alpha=0.1, fill="grey90", col="white") +
  # geom_line(method="lm", alpha=0.5, size=0.25, formula= y ~ sin(x/24*2*pi) + cos(x/24*2*pi), se=FALSE) +
  geom_line(stat="smooth", method="lm", alpha=0.3, size=0.25, formula= y ~ sin(x/24*2*pi) + cos(x/24*2*pi), se=FALSE) +
  # facet_wrap(~variable, nrow=2) + 
  scale_color_manual(values=c(rep("black", length(res_list)))) +
  geom_point(data= sample_points1, aes(x=zt, y=value), col=colorpalette[6], size=0.75,alpha=0.75) +
  geom_point(data= sample_points2, aes(x=zt, y=value), col=colorpalette[10], size=0.75,alpha=0.75) +
  geom_point(data= sample_points3, aes(x=zt, y=value), col=colorpalette[14], size=0.75,alpha=0.75) +
  geom_point(data= sample_points4, aes(x=zt, y=value), col=colorpalette[18], size=0.75,alpha=0.75) +
  geom_point(data= sample_points5, aes(x=zt, y=value), col="darkred", size=0.75, alpha=0.75)

```
### Figure 5f: boxplot

```{r fig.height= 3, fig.width=5}

sam_total %>%
  mutate(tmp = factor(tmp, levels=c("early\nmorning","morning","early\nafternoon","late\nafternoon","random\nduring work-hours"))) %>%
  ggplot(aes(x=tmp, y= value)) +
  geom_boxplot() +
  geom_dotplot(aes(fill=tmp),binaxis='y', stackdir='center', col = "white",stackratio=1, dotsize=1) + 
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "none") +
  ylab("cytosine modification density \n at sampling") + xlab("sampling window") +
  scale_fill_manual(values=c(colorpalette[6],colorpalette[10],colorpalette[14],colorpalette[18], "darkred")) +
  ylim(c(0,7)) 

```
### Figure 5f: statistical tests
```{r}

df1 <- sam_total %>%
  dcast(rep ~ tmp, value.var = "value", sum)

bres <- bartlett.test(data = sam_total, value ~ tmp)
ares <- aov(data = sam_total, value ~ tmp)
summary(ares)[[1]][["Pr(>F)"]][1]
# tres$p.value
bres$p.value
TukeyHSD(ares)

```

### Figure 5g: anti-phasic oscillations
```{r fig.height= 3, fig.width=5}

res_list <- vector('list', 50)
phase_shift = 2

for(i in 1:length(res_list)){
  hypothetical_amplitude <- 5
  hypo_amp <- sample(seq(hypothetical_amplitude/2, hypothetical_amplitude, hypothetical_amplitude/10),1)
  tmp2 <- data.frame(rep=i, zt = 0:48, 
                     control = coef2wave(amp=hypo_amp, int= 2.5, zts=0:48,
                                                 acro=sample(c(seq(23, 24, 0.5), seq(1, 5, 0.5)), 1))
                     # trough = coef2wave(amp=hypo_amp2, int= 2.5,
                     #                             acro=sample(c(seq(21, 24, 0.5), seq(0, 3, 0.5)), 1)),
                     # peak = coef2wave(amp=hypo_amp2, int= 2.5,
                     #                             acro=sample(c(seq(21, 24, 0.5), seq(0, 3, 0.5)), 1))
                     ) 
  tmp2
  tmp2$anti_phase <- c(tmp2$control[13:25], tmp2$control[1:12],tmp2$control[38:49], tmp2$control[26:37])
# length(tmp2$control)
  res_list[[i]] <- tmp2
  }

res <- do.call(rbind, res_list)

res <- res %>%
  melt(., id.var = c("rep","zt")) 

graph1 <- res %>%
  ggplot(aes(x=zt, y=value, fill=as.factor(rep), col=as.factor(variable))) +
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "none") +
  ylab("normalized\ncytosine modification\ndensity (%)") + xlab("time of day") +
  scale_x_continuous(breaks=seq(0,48, 6), labels = c("12AM","6AM", "12PM","6PM","12AM","6AM", "12PM","6PM", "12AM")) +
  annotate("rect", xmin=12,xmax=18,ymin=-Inf,ymax=Inf, alpha=0.5, fill="grey75", col="white") +
  annotate("rect", xmin=30,xmax=36,ymin=-Inf,ymax=Inf, alpha=0.5, fill=colorpalette[19], col="white") +
  geom_smooth(method="lm", size=0.25, formula= y ~ sin(x/24*2*pi) + cos(x/24*2*pi), se=FALSE) +
  # geom_hline(yintercept=2.5, col="red") +
  # facet_wrap(~variable, nrow=1) +
  scale_color_manual(values=c(colorpalette[6], colorpalette[15]))
  # facet_wrap(~variable) + scale_color_manual(values=c(viridis::cividis(20))) 

graph1

### Save output
# saveRDS(res, "output/antiphase.RDS")

### Load published data
res <- readRDS("input/antiphase.RDS")

```

### Figure 5g: sampling anti-phasic oscillations
```{r fig.height= 3, fig.width=5}
colorpalette = viridis::viridis(20)

sample_points1 <- res %>%
    filter(variable !="healthy_cotwins") %>%
  filter( zt >= 12, zt <= 18 ) %>%
  dcast(rep + zt ~ variable, value.var = "value", sum) %>%
  group_by(rep) %>%
  sample_n(., 1) %>%
  melt(., id.var = c("rep","zt")) %>%
  mutate(tmp ="afternoon")

sample_points2 <- res %>%
    filter(variable !="healthy_cotwins") %>%
  filter( zt >= 30, zt < 36 ) %>%
  dcast(rep + zt ~ variable, value.var = "value", sum) %>%
  group_by(rep) %>%
  sample_n(., 1) %>%
  melt(., id.var = c("rep","zt")) %>%
  mutate(tmp ="morning")

graph1 +
  geom_point(data= sample_points1, aes(x=zt, y=value, col=variable), size=0.5, alpha=0.75) +
  geom_point(data= sample_points2, aes(x=zt, y=value, col=variable), size=0.5, alpha=0.75) +
  ylim(-1, 6)

```

### Figure 5g: sampled anti-phasic oscillations
```{r fig.height= 3, fig.width=3.5}

rbind(sample_points1, sample_points2) %>%
  mutate(tmp = factor(tmp, levels=c("afternoon","morning"))) %>%
  ggplot(aes(x=variable, y= value)) +
  geom_boxplot() +
  geom_dotplot(aes(fill=variable),binaxis='y', stackdir='center', col = "white",stackratio=1, dotsize=1) + 
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "none") +
  ylab("cytosine modification \n at work-hour sampling") + xlab("time of sampling") +
  facet_wrap(~tmp) +
  scale_fill_manual(values=c(colorpalette[6], colorpalette[15])) 

```


